Rationality, Evolution \& the Environment

### 2015-05-27 paper/paper.tex
restructuring after Paolo section 2 first drafts

\section{Faces of rationality}

VERSION 1

A central pivot in economics and decision theory is the definition of rationality as maximization of subjective preferences. The subjectivity of preferences plays a very important role. From an agent's standpoint, the amount of, say, money achieved does not necessarily coincide with the amount of subjective utility attained. The first to notice this difference between objective utility (goods, money, etc.) and subjective utility (preferences) was D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that subjective utility can be represented by an increasing concave function of objective utility. This can be considered the birth of modern decision theory.
Another important step was developed by Kahnemann and Tversky who introduced prospect theory in their famous paper from 1979. By means of experimental results they noticed that the decision maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent with respect to losses. Consequently, they conjectured that the subjective utility is better representable by an S-shaped utility function, concave in the positive domain of gains and convex in the negative domain of losses.
More recently, other possible definitions of the subjective utility function have been proposed and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin, etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas (e.g., Prisoner's dilemma, public goods game, etc.). By means of different subjective utility functions, it is then possible to describe/match the behavior observed in experiments and to explain it as rational (in the sense of maximizing the subjective utility). \\
With this approach it seems that every behavior, no matter how far from the maximization of objective utility, can be deemed as rational if we can only find a subjective utility that justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost) everything real is rational. The problem with this approach is that it seems that everything that is observed can be considered rational, given the adequate subjective utility. The notion of rationality almost collapses onto the observed behavior.

@@

In particular, in the following we mainly deal with two types of agents: players whose utility function coincides with the objective utility, and players whose subjective utility is defined in terms of regret. The justification for this choice is both historical and behavioral. Among the possible competitors of utility maximization, regret has always been in a prominent position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes 
\begin{quote}
By explicitly incorporating regret, expected utility theory not only becomes a better descriptive predictor but also may become a more convincing guide for prescribing behavior to decision makers.
\end{quote}

\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match and explain the main experimental results of prospect theory, and claim that
\begin{quote}

[...] we shall challenge the idea that the conventional axioms constitute the only acceptable basis for rational choice under uncertainty. We shall argue that it is no less rational to act in accordance with regret theory, and that conventional expected utility theory therefore represents an unnecessarily restrictive notion of rationality. \cite{loosug82}

\end{quote}

...

END VERSION 1

VERSION 2

@@

END VERSION 2

\begin{itemize}
  \item \todo[inline]{choose uniform consensual terminology for maximization of EU and regret minimization}
\item classical notion of rationality: maximization of expected utilities
  \begin{itemize}
  \item human choice behavior seems to deviate from classical rational choice in many ways
    \citep[most
    notably][]{TverskyKahnemann1974:Judgement-under,KahnemannTversky1979:Prospect-Theory}
  \end{itemize}
\item historical competitor: regret minimization
  \begin{itemize}
  \item introduced by Savage \& Niehans independently
  \item worked out (in parameterized form) by \citet{LoomesSugden1982:Regret-Theory:-} and
    argued to outperform prospect theory
  \item recently rediscovered by game theorists \citep{HalpernPass2012:Iterated-Regret} as a
    possible partial explanation of some puzzles (e.g., Travellers Dilemma)
  \end{itemize}
\item motivate why we are interested in regret minimization; e.g.:
  \begin{itemize}
  \item simple historical alternative; certain prima facie plausibility
  \item has both components of interest:
    \begin{itemize}
    \item epistemic component: probabilistic vs.~non-probabilistic uncertainty  (interesting
      because of the classical Ellsberg paradox)
    \item payoff component: veridical vs.~non-veridical preference representations (interesting
      because of a recent interest in the ``evolution of preference'')
    \item gives us intriguing results that nicely motivate the general approach of meta-games
    \end{itemize}
  \end{itemize}
\end{itemize}

>>>>

\section{Rationality \& subjective preference}

\subsection{What should be in this section?}

\myalert{main purpose of this section: introduce max EU and min regret with some background,
  definitions, simple example and reasons why we care to look at especially these}

\begin{itemize}
  \item \todo[inline]{choose uniform consensual terminology for maximization of EU and regret minimization}
\item classical notion of rationality: maximization of expected utilities
  \begin{itemize}
  \item human choice behavior seems to deviate from classical rational choice in many ways
    \citep[most
    notably][]{TverskyKahnemann1974:Judgement-under,KahnemannTversky1979:Prospect-Theory}
  \end{itemize}
\item historical competitor: regret minimization
  \begin{itemize}
  \item introduced by Savage \& Niehans independently
  \item worked out (in parameterized form) by \citet{LoomesSugden1982:Regret-Theory:-} and
    argued to outperform prospect theory
  \item recently rediscovered by game theorists \citep{HalpernPass2012:Iterated-Regret} as a
    possible partial explanation of some puzzles (e.g., Travellers Dilemma)
  \end{itemize}
\item definitions of max EU and min reg (with minimal formal notation); simple example where
  they differ
\item motivate why we are interested in regret minimization; e.g.:
  \begin{itemize}
  \item simple historical alternative; certain prima facie plausibility
  \item argued to be empirically adequate (Loomes \& Sugden, Bell)
  \item gives us intriguing results that nicely motivate the general approach of meta-games
  \item non-independence of choices also for non-human decision makers (cite)
  \end{itemize}
\item here we care about game problems, not solitary decision problems
\end{itemize}

\subsection{Version 1}

A pivotal element of economics and decision theory is the definition of rationality as
maximization of subjective preferences. The subjectivity of preferences plays a very important
role. From an agent's standpoint, the amount of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained. The first to notice this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that
subjective utility can be represented by an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

Another important step was developed by Kahnemann and Tversky who introduced prospect theory in
their famous paper from 1979. By means of experimental results they noticed that the decision
maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). By means of different subjective utility
functions, it is then possible to describe/match the behavior observed in experiments and to
explain it as rational (in the sense of maximizing the subjective utility).

With this approach it seems that every behavior, no matter how far from the maximization of
objective utility, can be deemed as rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost)
everything real is rational. The problem with this approach is that it seems that everything
that is observed can be considered rational, given the adequate subjective utility. The notion
of rationality almost collapses onto the observed behavior.

@@

In particular, in the following we mainly deal with two types of agents: players whose utility
function coincides with the objective utility, and players whose subjective utility is defined
in terms of regret. The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}

\subsection{Version 2}

@@
